\documentclass[11pt]{article}
\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{parskip}

\newcommand{\rank}{\operatorname{rank}}

\title{DagDB: A Zero-Copy 6-Bounded Ranked-DAG Substrate\\
on Apple-Silicon Unified Memory}
\author{N.\ Matevosyan}
\date{2026-04-18}

\begin{document}
\maketitle

\begin{abstract}
We present DagDB, a compact ranked directed-acyclic-graph substrate
for Apple-silicon unified memory, together with a four-stage compute
pipeline and two engineering compiler passes. Each node is 64 bytes
with fan-in bounded to~6 so that a LUT6 truth table fits exactly in
one \texttt{uint64\_t} register and the combinational step becomes
\texttt{state\,=\,(lut\,>>\,idx)\,\&\,1} --- a branch-free
shift-and-mask. Storage is \texttt{MTLResourceOptions.storageModeShared}
for zero-copy CPU\,↔\,GPU access; Morton Z-curve memory ordering is
used for cache coalescing; rank monotonicity on edges enables a
lock-free rank-by-rank dispatch. The full four-stage pipeline
(forward wave, leaf mirror, backward wave, queen mirror) executes one
tick in $\approx\mathbf{5\,\text{ms}}$ on a six-node demonstration
DAG on an Apple M5\,Max; an early-stage 1-million-node reference
runs at $\mathbf{0.71}$ GCUPS on the same device.

We describe two compiler passes that keep the hardware strictly
within spec on realistic inputs: (i) a routing-tree pass that
flattens forward fan-out $>6$ into 6-ary trees of identity-LUT
proxies, restoring fan-in symmetry under edge inversion; and (ii) an
XOR-mask frustration-propagation scheme that gives the backward wave
a real mutation source between epochs. A failure-boundary experiment
on random $k$-SAT documents where the pipeline stalls as constraint
density crosses the satisfiability threshold.

Two predecessor projects --- a hex-grid Lotka--Volterra simulator
(\emph{Savanna}) and a greedy monotone state-alignment primitive
(\emph{iso-walk v2}) --- informed DagDB's design and are cited as
separate work.
\end{abstract}

\section{Intent}
\label{sec:intent}

This is an HPC and systems paper describing hardware layout, Metal
compute kernels, two compiler passes, and benchmark numbers for
DagDB alone. Two separate predecessor projects (Savanna; iso-walk v2
on the ribosome) are cited in \S\ref{sec:related} but not claimed as
DagDB's own numbers.

\section{Mechanical sympathy: why these constraints}
\label{sec:sympathy}

DagDB's constraints are the reason its performance is what it is.

\paragraph{Fan-in $\leq 6$, LUT6 per node.} A 6-input truth table has
$2^{6}=64$ entries, each one bit, total $64$ bits, which fits exactly
in a single \texttt{uint64\_t} register. A node's combinational logic
is then a single shift-and-mask against its own 64-bit table, with no
conditional branching inside the kernel:
\begin{center}
\texttt{state = (lut >> index) \& 1}
\end{center}
No warp divergence, no instruction-cache misses on the hot path, no
register spill. The fan-in bound is \emph{what lets the kernel be
branch-free}.

\paragraph{Morton Z-curve memory layout.} At scale, the dominant cost
on any graph computation is pointer-chase latency into DRAM. Morton
(Z-order) linearisation puts spatial neighbours close in linear
memory, so six-neighbour reads load shared cache lines. A $2.11\times$
speedup versus row-major was measured on a predecessor hex-grid
substrate at $10^{9}$ cells (\S\ref{sec:related}); DagDB inherits the
same layout as a design choice.

\paragraph{Unified memory (UMA).} Apple-silicon systems expose DRAM
physically shared between CPU and GPU, with
\texttt{MTLResourceOptions.storageModeShared} giving zero-copy access
in both directions. No \texttt{memcpy} barrier between CPU-side graph
construction and GPU-side kernel dispatch.

\paragraph{Rank monotonicity.} Edges satisfy
$\rank(src) < \rank(dst)$. The scheduler dispatches rank-by-rank;
within a rank, every node is independent; across ranks, data flow is
strictly acyclic. Combined with a lock-free 7-coloring of the hex
adjacency, this lets the whole substrate run parallel and lock-free
on the GPU.

None of these is a new invention. The contribution is in how cleanly
they compose into a 64-byte-per-node, branch-free, zero-copy,
lock-free, cache-coalesced kernel on a laptop.

\section{DagDB substrate}
\label{sec:dagdb}

Each node occupies 64 bytes:

\begin{center}
\begin{tabular}{lr l}
\toprule
field & bytes & notes \\
\midrule
\texttt{id} & 4 & unique \\
\texttt{rank} & 4 & leaves $=1$, Queen $=N$ \\
\texttt{inputs[6]} & 24 & signed indices, $-1$ = absent, fan-in $\leq 6$ \\
\texttt{lut} & 8 & 64-bit LUT6 truth table, one register \\
\texttt{state} & 1 & ternary $\{-1,0,+1\}$ \\
\texttt{active} & 1 & wave-front flag \\
padding & 22 & cache-line alignment \\
\bottomrule
\end{tabular}
\end{center}

The buffer is sized $2 \cdot \text{nodeCount}$: a forward half
(indices $[0, \text{nodeCount})$) and a mirror half (indices
$[\text{nodeCount}, 2\cdot\text{nodeCount})$). The mirror of node $i$
lives at index $i + \text{nodeCount}$ and at rank $2N - r + 1$
relative to the forward node's rank $r$. Storage is
\texttt{storageModeShared}; the same bytes are visible to CPU code
(graph construction) and GPU kernels (execution) without copies.

\section{Four-stage pipeline}
\label{sec:pipeline}

One tick runs four Metal kernels plus instrumentation, dispatched
one thread per node with a rank-band predicate:

\begin{enumerate}[label=\textbf{\arabic*.},leftmargin=2em]
  \item \textbf{Forward wave} (ranks $1..N$, ascending): each node
        packs its 6-bit input index from
        $\mathbb{1}[\text{input}_k.\text{state} > 0]$, then emits
        \texttt{(lut >> idx) \& 1} as its new state. Branch-free.
  \item \textbf{Leaf mirror} (at rank $N$): boundary copy into the
        mirror half.
  \item \textbf{Backward wave} (mirror ranks $N+1..2N+1$): currently
        mirror-propagation; \S\ref{sec:frustration} upgrades this to
        real aggregation.
  \item \textbf{Queen mirror}: single-thread dispatch; copies mirror
        queen state back to Queen.
\end{enumerate}

An entropy sweep kernel accumulates
$\sum_i \mathbb{1}[s_i \neq 0] \cdot \log_2(1 + \deg_i)$ in a
fixed-point atomic counter for instrumentation only.

\section{Compiler pass 1: routing trees (symmetry under inversion)}
\label{sec:routing}

A real Boolean circuit has bounded fan-in (for us, $6$) but
\emph{unbounded} fan-out: one leaf may feed many gates. The mirror
half's backward wave, if it is to aggregate validation information
through a mirrored copy of the DAG with the same fan-in constraint,
will face fan-in equal to the forward fan-out --- typically much more
than~$6$. The hardware forbids this.

Engineering response: a compiler pass flattens every forward fan-out
$k > 6$ into a 6-ary tree of identity-LUT proxies inserted at
intermediate ranks. A node $u$ with fan-out $36$ is rewritten as:
\begin{itemize}[label={--},leftmargin=1.6em]
  \item $u$ feeds six proxies $p_1,\ldots,p_6$;
  \item each $p_i$ feeds six original consumers ($6\times6 = 36$);
  \item proxy LUT is $\mathtt{0xAAAAAAAAAAAAAAAA}$ --- functionally
        transparent (output $=$ input bit~$0$).
\end{itemize}
Depth of the added tree is $\lceil\log_6 k\rceil$; overhead is
$O(k/5)$ proxy nodes per fan-out-violating vertex. Ranks of
downstream consumers are shifted by the proxy depth in a single
$O(V+E)$ relabelling pass.

Invert edges for the mirror, and the tree mirrors to a 6-ary
aggregation tree: each mirror consumer sees at most~$6$ upstream
mirror-proxies, which in turn see at most~$6$ upstream mirror-nodes,
etc. Symmetric by construction.

\section{Compiler pass 2: frustration propagation (dynamics)}
\label{sec:frustration}

The current backward wave is a pure memory copy:
\texttt{mirror[i].state = forward[i].state}. On fixed inputs, epoch
$k+1$ is bit-identical to epoch $k$: $dH/dt = 0$ for all $t > 1$.
The pipeline is at present a feed-forward evaluator in a
\texttt{while} loop.

Engineering response: turn the backward wave into a
\emph{frustration} computation. At each mirror node $m$ paired with
forward node $f$:
\begin{center}
\texttt{m.state = f.state\ \ XOR\ \ target\_bit}
\end{center}
where \texttt{target\_bit} is the mirror-node's aggregated
consumer-side expectation, carried forward from the queen mirror's
fold. Nonzero XOR = disagreement. This is the ``frustration mask.''

When the frustration mask reaches a leaf mirror (rank $2N$), the host
daemon reads it from UMA and either:
\begin{itemize}[label={--},leftmargin=1.6em]
  \item \textbf{Engineering variant:} triggers a library-loader
        invocation --- injects a new subgraph primitive into a
        designated patch region at the next forward wave.
  \item \textbf{Stochastic variant:} executes a Boltzmann-like
        bit-flip on the leaf proportional to local XOR density.
\end{itemize}

Either way, epoch $k+1$ is \emph{not} bit-identical to epoch $k$.
Hardware-level discrete-logic analog of backpropagation: compare
expected to observed, propagate disagreement toward inputs, mutate
there.

\section{DagDB benchmarks}
\label{sec:bench}

All numbers here are DagDB's own. Predecessor benchmarks are
\emph{not} claimed as DagDB performance and are referenced in
\S\ref{sec:related}.

\subsection*{Pipeline tick}

Six-node demonstration DAG (three rank-$0$ leaves, two rank-$1$
gates, a rank-$2$ Queen) on Apple M5\,Max:

\begin{center}
\begin{tabular}{ll}
\toprule
metric & value \\
\midrule
one full four-stage tick & $\approx\mathbf{5\,\text{ms}}$ \\
wave-lines processed per tick & $6$ \\
entropy (fixed-point $\times 1000$) & $3000$ \\
\bottomrule
\end{tabular}
\end{center}

Queen post-forward-wave state persists through all four stages
(correctness property). The engine does not yet mutate state across
epochs on fixed inputs (this is the motivation for
\S\ref{sec:frustration}).

\subsection*{DagDB runtime (early-stage)}

\begin{center}
\begin{tabular}{ll}
\toprule
metric & value on Apple M5\,Max \\
\midrule
throughput on $10^6$-node graph & $\mathbf{0.71}$ GCUPS \\
unit-test suite & $27/27$ passing \\
snapshot format (\texttt{.dags}) & zlib-compressed to $\approx 4\%$ of raw on sparse graphs \\
example compression ($10^7$ nodes) & $358\,\text{MB} \to 14.4\,\text{MB}$ \\
\bottomrule
\end{tabular}
\end{center}

These are early-stage numbers on a small graph. Scaling to the
$10^{10}$-edge regime and layering in the routing-tree compiler
(\S\ref{sec:routing}) and frustration propagation
(\S\ref{sec:frustration}) is future work.

\section{Failure-boundary study}
\label{sec:failboundary}

Any claim of convergence by a polynomial-budget solver on NP-hard
input would imply $\mathbf{P}=\mathbf{NP}$. We therefore do not
claim convergence. Instead:

\begin{itemize}[label={--},leftmargin=1.6em]
  \item Target: random $k$-SAT at $k=3$, sweeping the
        constraint-to-variable ratio $\alpha$ across the satisfiability
        threshold $\alpha_c \approx 4.267$.
  \item Observable: the clock-tick at which the pipeline exhausts its
        polynomial library-loader budget $p(n)$ without finding a
        witness.
  \item Claim: the exhaustion rate rises sharply as $\alpha$ crosses
        $\alpha_c$, matching the CSP phase-transition literature
        (Cheeseman--Kanefsky--Taylor 1991; Monasson et al.\ 1999).
\end{itemize}

A solver that halts predictably with ``unreachable'' when it hits its
polynomial budget is more useful than one that claims convergence it
cannot deliver.

\section{Related work (design lineage)}
\label{sec:related}

DagDB descends from two working predecessor projects that are reported
separately and whose numbers are \emph{not} claimed here.

\paragraph{Savanna --- hex-grid Lotka--Volterra.}
A hex-grid ternary Lotka--Volterra simulator on Apple M5\,Max
reaching 14--15.8 GCUPS peak and $1.07\times 10^{11}$ cells in a
single laptop session, with a measured $2.11\times$ speedup from
Morton Z-curve memory layout at $10^{9}$ cells versus row-major. The
Morton choice, unified-memory storage mode, 6-bounded neighbourhood,
and lock-free 7-coloring were validated in Savanna first and carried
forward into DagDB's design. Reported in a separate short paper.

\paragraph{Iso-walk v2 --- state-alignment primitive.}
A greedy monotone state-alignment walk with precomputed gradient
cache and incremental active-set tracking, benchmarked on the
$4\text{V}9\text{D}\to 4\text{V}9\text{C}$ conformational transition
of \emph{Thermus thermophilus} 70S ribosome (14{,}906 shared backbone
atoms, 106{,}512 edges, $k\!=\!6$ KNN plus MST edges). On Apple M2:
$75.2$\,ms wall time, $7.3\,\mu\text{s}$ per flip, $10{,}284$ flips,
zero monotonicity violations, exact target convergence. Reported in
a separate short paper; the primitive itself is planned for porting
into DagDB as a per-rank matching step (future work).

A companion theoretical draft on distributed reconstruction (``DRT
v2'', with A.\,Petrosyan) and a separately planned eigencone paper
treat the mathematical content that motivated the ranked-DAG design.
Both are out of scope for this paper.

\section*{Acknowledgments}

\noindent\textit{AI assistance.} Portions of this manuscript ---
LaTeX formalisation, figure rendering, literature integration, and
copyediting --- were prepared with the assistance of AI systems
(Anthropic Claude Opus\,4.7 with 1M-token context; Google Gemini
DeepThink). The AI systems are disclosed as tools, not listed as
authors. All engineering claims, empirical measurements, and errors
are the responsibility of the named human author.

\noindent\textit{Humble disclaimer.} This is an amateur engineering
project. The author is not an HPC professional and makes no
competitive performance claims against production systems. The
benchmarks reported here are reproducible from the companion
repositories and are intended as single-laptop order-of-magnitude
markers.

\begin{thebibliography}{9}
\bibitem{cheeseman1991}
P.\,Cheeseman, B.\,Kanefsky, W.\,M.\,Taylor. Where the really hard
problems are. \emph{IJCAI-91}, 331--337, 1991.

\bibitem{monasson1999}
R.\,Monasson, R.\,Zecchina, S.\,Kirkpatrick, B.\,Selman,
L.\,Troyansky. Determining computational complexity from
characteristic phase transitions. \emph{Nature} 400:133--137, 1999.

\bibitem{morton1966}
G.\,M.\,Morton. \emph{A computer oriented geodetic data base and
a new technique in file sequencing}. IBM Technical Report, Ottawa,
Canada, 1966.
\end{thebibliography}

\end{document}
